---
title: "oak_data_cleaning"
author: "Kelsey Hauf"
format: html
editor: visual
---

## Load Libraries + Data

```{r}
#load libaraies
library(tidyverse)
library(dplyr)
library(lattice)
library(nlme)
library(car)
library(ggplot2)
```

## cleaning up oak data

```{r}

#size of the seedlings when planted
initial_size <- read_csv('Oak_Seedlings_Initial_size.csv')

initial_size <- initial_size |>
  select(-`Page`)

#seedling height, # of leaves, for winter + spring
seedling_data <- read_csv('Oak_Seedlings_Size.csv')

#spring acorn germination 
spring24germ <- read_csv('Spring24Germination.csv')

#spring soil moisture data
springSM <- read_csv('SpringSoilMoistureORPOY.csv')

seedling_acorn_data_spring <- read_csv('2024 Spring Seedling & Acorn Height and Survival ORPOY - Spring Data.csv')

#making the cloumn names match the preexisting data sets 
seedling_acorn_data_spring <- seedling_acorn_data_spring |>
  select(-`Data Enterer Initials`, -`Notes`, -`Page Number`, -`Date`) |>
  rename('Block' = 'Block Number') |>
  rename('Tag' = 'Tag Number') |>
  rename('LifeStage' = 'Acorn or seedling?') |>
  rename('Height' = 'Seedling Height') |>
  rename('GerminatedAcorns' = 'Number of Acorns That Germinated') |>
  rename('Leaves' = '# Leaves on Seedling') |>
  rename('Salvia' = 'Is Salvia apiana alive') 




```

## Create a data set that has the heights across seasons to compare them

```{r}

# Filter the initial seedling heights data
initial_heights <- initial_size |> 
  filter(LifeStage == 'Seedling', Treatment != 'NR') |> 
  select(Tag, Treatment, Height, Block) |>  
  mutate(Season = 'Initial') |>
  mutate(Lifestage = 'Seedling')

# Filter the spring seedling heights data
spring_heights <- seedling_acorn_data_spring |> 
  filter(LifeStage == 'Seedling') |> 
  select(Tag, Treatment, Height, Block) |> 
  mutate(Season = 'Spring') |>
  mutate(Lifestage = 'Seedling')


# Combine the two datasets
seedling_height_data <- bind_rows(initial_heights, spring_heights)

# View the combined data
#print(seedling_height_data)
  

```

# Visualizations - Heights

Create a graph that compares the overall heights

```{r}
seedling_height_data |>

  ggplot() +
  aes(x = Season , y = Height, fill = Season) +
  geom_col(position = position_dodge())

```

Create a graph that compares the change in heights by treatment

```{r}
seedling_height_data |>

  ggplot() +
  aes(x = Treatment , y = Height, fill = Season) +
  labs(x = "Treatment", y = "Height", fill = "Season") +  # Add labels
  geom_col(position = position_dodge())


```

Create a graph that compares the change in heights by block

```{r}
seedling_height_data |>

  ggplot() +
  aes(x = Block , y = Height, fill = Season) +
  labs(x = "Block", y = "Height", fill = "Season") +  # Add labels
  geom_col(position = position_dodge())
```

## Soil Moisture

```{r}

#spring 2024 soil moisture 

#If Treatment was previously a character or numeric type, this line changes it to a factor, which can be important for plotting or running statistical models that require categorical variables.

springSM$Treatment <- as.factor(springSM$Treatment)
springSM$Block <- as.factor(springSM$Block)


#plot the data:
plot(springSM$Treatment, springSM$VWC1, 
     xlab = "Planting Location", 
     ylab="Volumetric Water Content")



#let's make a new variable, the average of VWC

# Replace NAs with the mean of each VWC column
springSM <- springSM |>
  mutate(across(c(VWC1, VWC2, VWC3), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))

#-------------------------------------------------------------------------------

# Calculate MeanVWC
MeanVWC <- springSM |>
  mutate(MeanVWC = (VWC1 + VWC2 + VWC3) / 3) |>
  select(MeanVWC, Treatment, Block)


#------------------------------------------------------------------------------


# Create the bar plot showing mean vwc in each treatment 
ggplot(data = MeanVWC, aes(x = Treatment, y = MeanVWC)) +
  geom_bar(stat = "identity", position = position_dodge()) +  # Ensure you include stat = "identity"
  labs(x = "Planting Location", y = "Mean Volumetric Water Content")   # Add labels
  
# Create box and whisker graph showing mean vwc in each treatment 
plot(springSM$Treatment, springSM$MeanVWC, 
     xlab = "Planting Location", 
     ylab="Volumetric Water Content")

#----------------------------------------------------------------------------

#lets do the same for the blocks 

# Create the bar plot showing mean vwc in each block 
ggplot(data = MeanVWC, aes(x = Block, y = MeanVWC)) +
  geom_bar(stat = "identity", position = position_dodge()) +  # Ensure you include stat = "identity"
  labs(x = "Planting Location", y = "Mean Volumetric Water Content")   # Add labels
  

plot(springSM$Block, springSM$MeanVWC, 
     xlab="Experimental Block", 
     ylab="Volumetric Water Content")



str(springSM)
histogram(springSM$MeanVWC)



soil.aov <- aov(MeanVWC ~ Treatment, data=springSM)
summary(soil.aov)

soil.aov2 <- glm(MeanVWC ~ Treatment + Block, data=springSM)
Anova(soil.aov2)

#run a mixed model anova to determine whether soil moisture varied 
#depending on plot type while controlling for random Block effects

#soil.aov3 <- lme(MeanVWC~Treatment, random = ~1|Block, data=springSM)
#print(summary(soil.aov3))
#anova(soil.aov3)
```

## Germination

july 1st - june 30th

average blocks separate by treatment - effect on treatment

NR compare

Soil

do the soils vary

do they vary based on micro habitat -\> treatment

block is a random effect -\> how soils vary based on treatment

block as a fixed do soils vary across site -\> notice yes

controlling for the variation how do they vary across blocks\
