---
title: "oak_data_EDA"
author: "Kelsey Hauf"
format: html
editor: visual
---

# Load Libraries + Data

```{r}
#load libaraies
library(tidyverse)
library(dplyr)
library(lattice)
library(nlme)
library(car)
library(ggplot2)
install.packages("DT")
library(DT)
library(GGally)
library(scatterplot3d)

```

## Load + Clean Oak Data

```{r}

#Winter (initial) seedling data 

#size of the seedlings when planted
initial_size <- read_csv('Oak_Seedlings_Initial_size.csv')

initial_size <- initial_size |>
  select(-`Page`)


#spring 2024 soil moisture data
springSM <- read_csv('SpringSoilMoistureORPOY.csv')


# seedling height, # of leaves, and acorn germination for spring 2024
spring24 <- read_csv('2024 Spring Seedling & Acorn Height and Survival ORPOY - Spring Data.csv')

#making the cloumn names match the preexisting data sets 
spring24 <- spring24 |>
  select(-`Data Enterer Initials`, -`Notes`, -`Page Number`, -`Date`) |>
  rename('Block' = 'Block Number') |>
  rename('Tag' = 'Tag Number') |>
  rename('LifeStage' = 'Acorn or seedling?') |>
  rename('Height' = 'Seedling Height') |>
  rename('GerminatedAcorns' = 'Number of Acorns That Germinated') |>
  rename('Leaves' = '# Leaves on Seedling') |>
  rename('Salvia' = 'Is Salvia apiana alive') 

#summer soil moisture 
summerSM <- read_csv('2024 Summer Soil Moisture ORPOY - Summer24.csv')

# seedling height, # of leaves, and acorn germination for summer
summer24 <- read_csv('2024 Summer Seedling & Acorn Height and Survival ORPOY.xlsx - Summer Data.csv')

#making the cloumn names match the preexisting data sets 
summer24 <- summer24 |>
  select(-`Data Enterer Initials`, -`Notes`, -`Page Number`, -`Date`) |>
  rename('Block' = 'Block Number') |>
  rename('Tag' = 'Tag Number') |>
  rename('LifeStage' = 'Acorn or seedling?') |>
  rename('Height' = 'Seedling Height') |>
  rename('GerminatedAcorns' = 'Number of Acorns That Germinated') |>
  rename('Leaves' = '# Leaves on Seedling') |>
  rename('Salvia' = 'Is Salvia apiana alive') 


# seedling height, # of leaves, and acorn germination for Fall 2024
fall24 <- read_csv('2024 Fall Seedling & Acorn Height and Survival ORPOY - Sheet1.csv')

#making the cloumn names match the preexisting data sets 
fall24 <- fall24 |>
  select(-`Data Enterer Initials`, -`Page Number`, -`Date`, -`Data Entered`, -`QAQC`) |>
  rename('Block' = 'Block Number') |>
  rename('Tag' = 'Tag Number') |>
  rename('LifeStage' = 'Acorn or seedling?') |>
  rename('Height' = 'Seedling Height (cm)') |>
  rename('GerminatedAcorns' = 'Number of Acorns That Germinated') |>
  rename('Leaves' = '# Leaves on Seedling') |>
  rename('Salvia' = 'Is Salvia apiana alive') 

#Fall 2024 soil moisture data
fallSM <- read_csv('2024 Fall Soil Moisture ORPOY - Sheet1.csv')

```

##   Number of Living Seedlings by Treatment and Block

```{r}
# Filter for living seedlings based on parameters
living_seedlings <- fall24 |>
  filter(LifeStage == "Seedling", Height > 1, Leaves > 1)

# Count living seedlings per block or overall
seedling_counts <- living_seedlings |>
  count(Block)

# Create a bar graph showing living seedlings per block
ggplot(seedling_counts, aes(x = Block, y = n)) +
  geom_bar(stat = "identity", fill = "forestgreen") +
  labs(
    title = "Number of Living Seedlings in Fall 2024",
    x = "Block",
    y = "Number of Living Seedlings"
  ) +
  theme_minimal()

# Total tally of seedlings
total_seedlings <- fall24 |>
  filter(LifeStage == "Seedling") |>
  nrow()

# Tally of alive seedlings
alive_seedlings <- fall24 |>
  filter(LifeStage == "Seedling", Height > 1, Leaves > 1) |>
  nrow()

# Print the results
cat("Total number of seedlings:", total_seedlings, "\n")
cat("Number of alive seedlings:", alive_seedlings, "\n")

# Count living seedlings by block and treatment
living_seedlings_by_treatment <- fall24 |>
  filter(LifeStage == "Seedling", Height > 1, Leaves > 1) |>
  count(Block, Treatment)

# Create a faceted bar graph
ggplot(living_seedlings_by_treatment, aes(x = Treatment, y = n, fill = Treatment)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ Block) +
  labs(
    title = "Number of Living Seedlings by Treatment and Block",
    x = "Treatment",
    y = "Number of Living Seedlings",
    fill = "Treatment",
    caption = paste("Total seedlings that survived across all blocks:", alive_seedlings,"\n",
      "Out of:", total_seedlings)
   
  ) #+
  #theme_minimal() #+
 # theme(
 #   strip.text = element_text(size = 12, face = "bold"),
 #   axis.text.x = element_text(angle = 45, hjust = 1)
  #)



#change block number to a char to show block number 
```

## Number of germinated acorns by block and treatment

```{r}
# Summarizing the number of germinated acorns by block and treatment
fall24_summary <- fall24 %>%
  group_by(Block, Treatment) %>%
  summarise(Total_Germinated = sum(GerminatedAcorns, na.rm = TRUE))

# Calculate total germinated acorns across all blocks
total_germinated <- sum(fall24_summary$Total_Germinated, na.rm = TRUE)

# Get the maximum value for y-axis scaling
max_value <- max(fall24_summary$Total_Germinated, na.rm = TRUE)

# Creating the faceted plot with a total count in the caption
ggplot(fall24_summary, aes(x = Treatment, y = Total_Germinated, fill = Treatment)) +
  geom_bar(stat = "identity") +
  facet_grid(~Block) +  # Facet by Block
  labs(
    title = "Number of Germinated Acorns by Treatment (Faceted by Block)",
    x = "Treatment",
    y = "Total Germinated Acorns",
    fill = "Treatment",
    caption = paste("Total germinated acorns across all blocks:", total_germinated)  # Add text under legend
  ) +
  scale_y_continuous(breaks = seq(1, max_value, by = 1)) +  # Set y-axis breaks
  #theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Create a data set that has the heights across seasons to compare them

```{r}

# Filter the initial seedling heights data and add a column that indicates what season its from
initial_heights <- initial_size |> 
  filter(LifeStage == 'Seedling', Treatment != 'NR') |> 
  select(Tag, Treatment, Height, Block) |>  
  mutate(Season = 'Initial24') |>
  mutate(Lifestage = 'Seedling')

# Filter the spring24 seedling heights data
spring_heights <- spring24 |> 
  filter(LifeStage == 'Seedling') |> 
  select(Tag, Treatment, Height, Block) |> 
  mutate(Season = 'Spring24') |>
  mutate(Lifestage = 'Seedling')


# Filter the summer24 seedling heights data
summer_heights <- summer24 |> 
  filter(LifeStage == 'Seedling') |> 
  select(Tag, Treatment, Height, Block) |> 
  mutate(Season = 'Summer24') |>
  mutate(Lifestage = 'Seedling')


# Filter the fall24 seedling heights data
fall_heights <- fall24 |> 
  filter(LifeStage == 'Seedling') |> 
  select(Tag, Treatment, Height, Block) |> 
  mutate(Season = 'Fall24') |>
  mutate(Lifestage = 'Seedling')



# Combine the  datasets
seedling_height_data <- bind_rows(initial_heights, spring_heights, summer_heights, fall_heights)

# View the combined data
print(seedling_height_data)

seedling_height_data <- seedling_height_data |>
  mutate(Season = factor(Season, levels = c("Initial24", "Spring24", "Summer24", "Fall24")))


```

### Height

Create a graph that compares the overall heights

```{r}
#take out NR treatment 
seedling_height_data <- seedling_height_data |> 
  filter(Treatment != 'NR') 
  

seedling_height_data |>

  ggplot() +
  aes(x = Season , y = Height, fill = Season) +
  geom_col(position = position_dodge())

# remember there is an extreme outlier in the 8th block where 1 plant had over 50 cm of growth 

```

#### Create a graph that compares the change in heights by treatment

```{r}
seedling_height_data |>

  ggplot() +
  aes(x = Treatment , y = Height, fill = Season) +
  labs(x = "Treatment", y = "Height", fill = "Season") +  # Add labels
  geom_col(position = position_dodge())


```

#### Create a graph that compares the change in heights by block

```{r}
seedling_height_data |>

  ggplot() +
  aes(x = Block , y = Height, fill = Season) +
  labs(x = "Block", y = "Height", fill = "Season") +  # Add labels
  geom_col(position = position_dodge())
```

## Growth Rate

calculate and compare the growth rates between the seasons

```{r}
# Pivot wider to have heights for each season in separate columns
seedling_height_wide <- seedling_height_data |>
  pivot_wider(names_from = Season, values_from = Height)

# Calculate growth rates
seedling_height_wide <- seedling_height_wide |>
  mutate(
    Growth_Initial_to_Spring = (Spring24 - Initial24) / Initial24,
    Growth_Spring_to_Summer = (Summer24 - Spring24) / Spring24,
    Growth_Summer_to_Fall = (Fall24 - Summer24) / Summer24,
  )

# Pivot longer to prepare for visualization
growth_rate_data <- seedling_height_wide |>
  select(Tag, Treatment, Block, Growth_Initial_to_Spring, Growth_Spring_to_Summer, Growth_Summer_to_Fall) |>
  pivot_longer(
    cols = starts_with("Growth"),
    names_to = "Growth_Period",
    values_to = "Growth_Rate"
  )

# Visualization: Compare growth rates by treatment
growth_rate_data |>
  ggplot() +
  aes(x = Treatment, y = Growth_Rate, fill = Growth_Period) +
  geom_col(position = position_dodge()) +
  labs(
    x = "Treatment",
    y = "Growth Rate",
    fill = "Growth Period",
    title = "Comparison of Growth Rates by Treatment"
  )
```

### Summarize growth rate patterns across treatments/seasons

```{r}

growth_rate_data |>
  group_by(Treatment, Growth_Period) |>
  summarize(
    Mean_Growth = mean(Growth_Rate, na.rm = TRUE),
    SD_Growth = sd(Growth_Rate, na.rm = TRUE),
    .groups = "drop"
  )

#plot it 
ggplot(growth_rate_data, 
       aes(x = Treatment, 
           y = Growth_Rate, 
           fill = Growth_Period)) +
  
  geom_boxplot() +
  labs(x = "Treatment", 
       y = "Growth Rate", 
       fill = "Growth Period")


```

### Statistical test

Determine if growth rates differ significantly between treatments

```{r}
#remove any NAs
growth_rate_data <- growth_rate_data |>
  filter(!is.na(Growth_Rate) & is.finite(Growth_Rate))

#ANOVA for treatments
anova_result <- aov(Growth_Rate ~ Treatment, data = growth_rate_data)
summary(anova_result)

#as of 11/21 The treatments did not significantly affect growth rates.
#as of 1/23 The treatments did not significantly affect growth rates.


```

### **Temporal Analysis**

Analyze how growth rates evolve over time and compare trajectories between treatments.

```{r}
ggplot(seedling_height_data, 
       aes(x = Season, 
           y = Height, 
           color = Treatment, 
           group = Tag)) +
  
  geom_line() +
  labs(x = "Season", 
       y = "Height", 
       color = "Treatment", 
       title = "Growth Rate change over time by Block") +
 
   facet_wrap(~ Block)  # Create a separate plot for each Block

```

### what about temperature? does it have an effect on growth rate

```{r}
#load in temp data 
HOBOData24 <- readRDS("HOBOData24.rds")

# Remove rows with NA in Growth_Rate or Avg_Temperature
#final_data_clean <- final_data |>
#  filter(!is.na(Growth_Rate) & !is.na(Avg_Temperature))

# Convert Block and Treatment to character 
growth_rate_data$Block <- as.character(growth_rate_data$Block)
HOBOData24$Block <- as.character(HOBOData24$Block)
growth_rate_data$Treatment <- as.character(growth_rate_data$Treatment)
HOBOData24$Treatment <- as.character(HOBOData24$Treatment)


# Merging the growth data with temperature data
merged_data <- merge(growth_rate_data, HOBOData24, by = c("Block", "Treatment"))

# Calculate average temperature by Block and Treatment
agg_temp_data <- HOBOData24 |>
  group_by(Block, Treatment) |>
  summarise(Avg_Temperature = mean(Temp, na.rm = TRUE))

# Merging the temperature data with the growth rate data
final_data <- merge(growth_rate_data, agg_temp_data, by = c("Block", "Treatment"))

# Linear model to test the effect of temperature on growth rate
growth_rate_temp_model <- lm(Growth_Rate ~ Avg_Temperature, data = final_data)

# Summary of the model
summary(growth_rate_temp_model)

# Model diagnostics
plot(growth_rate_temp_model)


#Temperature is a significant predictor of growth rate yay!


#visualize it 

ggplot(final_data, aes(x = Avg_Temperature, y = Growth_Rate)) +
  geom_point(aes(color = Block, shape = Treatment), position = position_jitter(width = 0.1, height = 0), alpha = 0.7) +  # Jittering and adding Treatment shape
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Add linear regression line
  labs(x = "Average Temperature", y = "Growth Rate", title = "Growth Rate vs Average Temperature") +
  theme_minimal() +
  scale_color_discrete(name = "Block") +  # Legend for Block
  scale_shape_manual(values = c(15, 17, 18, 19))  # Manually specify shapes for Treatment (can adjust the values for different shapes)






#------------------------------------------------------------------------
#Graveyard

# Including treatment as additional predictor
#growth_rate_temp_model2 <- lm(Growth_Rate ~ Avg_Temperature + Treatment, data = final_data)

# Summary of the updated model
#summary(growth_rate_temp_model2)

```

Now lets see if soil moisture has an effect on growth rate

```{r}
# Ensure 'Treatment' and 'Block' are factors for soil moisture dataset
springSM$Treatment <- as.factor(springSM$Treatment)
springSM$Block <- as.factor(springSM$Block)

fallSM$Treatment <- as.factor(fallSM$Treatment)
fallSM$Block <- as.factor(fallSM$Block)

#add a season column 

springSM <- springSM |>
  mutate(Season = "Spring 2024")

fallSM <- fallSM |>
  mutate(Season = "Fall 2024")



# Calculate the mean soil moisture (MeanVWC) as before
springSM <- springSM |>
  mutate(across(c(VWC1, VWC2, VWC3), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .))) |>
  mutate(MeanVWC = (VWC1 + VWC2 + VWC3) / 3)


fallSM <- fallSM |>
  mutate(across(c(VWC1, VWC2, VWC3), ~ as.numeric(.))) |>  # Convert to numeric
  mutate(across(c(VWC1, VWC2, VWC3), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .))) |>  # Replace NA with column mean
  mutate(MeanVWC = (VWC1 + VWC2 + VWC3) / 3)  # Calculate MeanVWC

# Ensure growth data also has 'Treatment' and 'Block' to enable merging
growth_rate_data$Treatment <- as.factor(growth_rate_data$Treatment)
growth_rate_data$Block <- as.factor(growth_rate_data$Block)



# Merge soil moisture and growth datasets by 'Treatment' and 'Block'
combinedSM_data <- merge(springSM, fallSM, by = c("Treatment", "Block"))
combinedSM_data <- merge(combinedSM_data, growth_rate_data, by = c("Treatment", "Block"))

# Quick look at combined dataset
head(combinedSM_data)

# Scatter plot of MeanVWC vs Growth Rate (Height or Leaves growth)
ggplot(combinedSM_data, aes(x = MeanVWC.y, y = Growth_Rate)) +
  geom_point(aes(color = Treatment), size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = "black") +
  labs(
    x = "Mean Spring 2024 Volumetric Water Content",
    y = "Growth Rate (Change in Height or Leaves)",
    title = "Effect of Soil Moisture on Growth Rate"
  ) 

# Linear regression to assess the relationship
model <- lm(Growth_Rate ~ MeanVWC.y, data = combinedSM_data)
summary(model)

# there is a statistically significant relationship between growth rate and VWC
```

long story short soil moisture does not have an effect on growth rate womp womp

## Leaf Data

### Initial Leaf Count EDA

First we want to prove statistically if there is a relationship between leaf count and height

```{r}
# Create new columns with log-transformed values for Height and Leaves - Log transformation ensures the data meets the assumptions required for valid statistical analyses
initial_size$lnHeight <- log(initial_size$Height + 1)
initial_size$lnLeaves <- log(initial_size$Leaves + 1)

# Histogram of the log-transformed Height - shows us if there is a normal distribution 
library(lattice)
histogram(initial_size$lnHeight)

# Scatter plot of lnHeight vs lnLeaves
plot(initial_size$lnHeight, initial_size$lnLeaves, 
     xlab = "Ln(Initial Height)", ylab = "Ln(Number of Leaves)")

# Linear model with lnHeight as the response and lnLeaves as the predictor - tells us if there is a strong relationship between height and leaves. spoiler there very much is
height.lvs.lm <- lm(lnHeight ~ lnLeaves, data = initial_size)

# Summary of the linear model
summary(height.lvs.lm)

# Diagnostic plots for the linear model - Ensures the model is appropriate for the data.
plot(height.lvs.lm)
```

There is a relationship which is great!

### Create a data set that has the leaf counts across seasons

```{r}

# Filter the spring24 leaf count data
spring_leaves <- spring24 |> 
  filter(LifeStage == 'Seedling', Treatment != 'NR') |> 
  select(Tag, Treatment, Leaves, Block) |> 
  mutate(Season = 'Spring') |>
  mutate(Lifestage = 'Seedling')


# Filter the summer24 leaf count data
summer_leaves <- summer24 |> 
  filter(LifeStage == 'Seedling', Treatment != 'NR') |> 
  select(Tag, Treatment, Leaves, Block) |> 
  mutate(Season = 'Summer') |>
  mutate(Lifestage = 'Seedling')


# Filter the fall24 leaf count data
fall_leaves <- fall24 |> 
  filter(LifeStage == 'Seedling', Treatment != 'NR') |> 
  select(Tag, Treatment, Leaves, Block) |> 
  mutate(Season = 'Fall') |>
  mutate(Lifestage = 'Seedling')


# Combine the  datasets
seedling_leaf_data <- bind_rows(spring_leaves, summer_leaves, fall_leaves)

# View the combined data
print(seedling_leaf_data)
  
```

### Visualizations

Leaf count across seasons

```{r}

seedling_leaf_data <- seedling_leaf_data |>
  mutate(Season = factor(Season, levels = c("Spring", "Summer", "Fall")))


seedling_leaf_data |>

  ggplot() +
  aes(x = Season , y = Leaves, fill = Season) +
  geom_col(position = position_dodge())

ggplot(seedling_leaf_data, aes(x = Season, y = Leaves, group = Tag, color = Tag)) +
  geom_line() +
  geom_point() +
  labs(x = "Season", y = "Leaf Count", title = "Change in Leaf Count Across Seasons") +
  theme_minimal()

ggplot(seedling_leaf_data, aes(x = Season, y = Leaves, group = Tag, color = Treatment)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ Block) +  # Or replace with Block or another grouping variable
  labs(x = "Season", y = "Leaf Count", title = "Change in Leaf Count Across Seasons by Treatment") +
  theme_minimal()


```

## Soil Moisture

```{r}

#spring 2024 soil moisture 

#If Treatment was previously a character or numeric type, this line changes it to a factor, which can be important for plotting or running statistical models that require categorical variables.

springSM$Treatment <- as.factor(springSM$Treatment)
springSM$Block <- as.factor(springSM$Block)


#plot the data:
plot(springSM$Treatment, springSM$VWC1, 
     xlab = "Planting Location", 
     ylab="Spring 2024 Volumetric Water Content")



#let's make a new variable, the average of VWC

# Replace NAs with the mean of each VWC column
springSM <- springSM |>
  mutate(across(c(VWC1, VWC2, VWC3), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))

#-------------------------------------------------------------------------------

# Calculate MeanVWC
MeanVWC <- springSM |>
  mutate(MeanVWC = (VWC1 + VWC2 + VWC3) / 3) |>
  select(MeanVWC, Treatment, Block)


#------------------------------------------------------------------------------


# Create the bar plot showing mean vwc in each treatment 
ggplot(data = MeanVWC, aes(x = Treatment, y = MeanVWC)) +
  geom_bar(stat = "identity", position = position_dodge()) +  # Ensure you include stat = "identity"
  labs(x = "Planting Location", y = "Mean Spring 2024 Volumetric Water Content")   # Add labels
  
# Create box and whisker graph showing mean vwc in each treatment 
plot(springSM$Treatment, springSM$MeanVWC, 
     xlab = "Planting Location", 
     ylab="Spring 2024 Volumetric Water Content")

#----------------------------------------------------------------------------

#lets do the same for the blocks 

# Create the bar plot showing mean vwc in each block 
ggplot(data = MeanVWC, aes(x = Block, y = MeanVWC)) +
  geom_bar(stat = "identity", position = position_dodge()) +  # Ensure you include stat = "identity"
  labs(x = "Planting Location", y = "Mean Spring 2024 Volumetric Water Content")   # Add labels
  

#plot(springSM$Block, springSM$MeanVWC, 
     #xlab="Experimental Block", 
     #ylab="Spring 2024 Volumetric Water Content")


#fall 2024 soil moisture 

#If Treatment was previously a character or numeric type, this line changes it to a factor, which can be important for plotting or running statistical models that require categorical variables.

fallSM$Treatment <- as.factor(fallSM$Treatment)
fallSM$Block <- as.factor(fallSM$Block)


#plot the data:
plot(fallSM$Treatment, fallSM$VWC1, 
     xlab = "Planting Location", 
     ylab="Fall 2024 Volumetric Water Content")



#let's make a new variable, the average of VWC

# Replace NAs with the mean of each VWC column
fallSM <- fallSM |>
  mutate(across(c(VWC1, VWC2, VWC3), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))

#-------------------------------------------------------------------------------

# Calculate MeanVWC
MeanVWC <- fallSM |>
  mutate(MeanVWC = (VWC1 + VWC2 + VWC3) / 3) |>
  select(MeanVWC, Treatment, Block)


#------------------------------------------------------------------------------


# Create the bar plot showing mean vwc in each treatment 
ggplot(data = MeanVWC, aes(x = Treatment, y = MeanVWC)) +
  geom_bar(stat = "identity", position = position_dodge()) +  # Ensure you include stat = "identity"
  labs(x = "Planting Location", y = "Mean Fall 2024 Volumetric Water Content")   # Add labels
  
# Create box and whisker graph showing mean vwc in each treatment 
plot(fallSM$Treatment, fallSM$MeanVWC, 
     xlab = "Planting Location", 
     ylab="Fall 2024 Volumetric Water Content")

#----------------------------------------------------------------------------

#lets do the same for the blocks 

# Create the bar plot showing mean vwc in each block 
ggplot(data = MeanVWC, aes(x = Block, y = MeanVWC)) +
  geom_bar(stat = "identity", position = position_dodge()) +  # Ensure you include stat = "identity"
  labs(x = "Planting Location", y = "Mean Fall 2024 Volumetric Water Content")   # Add labels
  
```

### Mean VWC by treatment and season 

```{r}

# Add a "Season" column to each dataset
springSM <- springSM |>
  mutate(Season = "Spring 2024")

fallSM <- fallSM |>
  mutate(Season = "Fall 2024")

fallSM <- fallSM |>
  mutate(across(c(PER1, PER2, PER3), ~ as.numeric(.))) #|>  # Convert to numeric
 # mutate(across(c(VWC1, VWC2, VWC3), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))   # Replace NA with column mean

# Combine the datasets
combinedSM_data <- bind_rows(springSM, fallSM)

# Calculate MeanVWC for the combined dataset
combinedSM_data <- combinedSM_data |>
  mutate(across(c(VWC1, VWC2, VWC3), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .))) |>
  mutate(MeanVWC = (VWC1 + VWC2 + VWC3) / 3)


ggplot(data = combinedSM_data, aes(x = Treatment, y = MeanVWC, fill = Season)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Planting Location", y = "Mean Volumetric Water Content", fill = "Season") +
  theme_minimal()


# uhhhhh not sure whats happening here? 
# log transform fixes everything 

# Log transform the VWC columns
combinedSMlog <- combinedSM_data |>
  mutate(across(c(VWC1, VWC2, VWC3), ~ log1p(.), .names = "log_{col}")) |>
  mutate(MeanVWC_Log = (log_VWC1 + log_VWC2 + log_VWC3) / 3)

ggplot(data = combinedSMlog, aes(x = Treatment, y = MeanVWC_Log, fill = Season)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Planting Location", y = "Log-Transformed Mean Volumetric Water Content", fill = "Season") +
  theme_minimal()
```

### Mean VWC by Block and season 

```{r}
ggplot(data = combinedSM_data, aes(x = Block, y = MeanVWC, fill = Season)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Experimental Block", y = "Mean Volumetric Water Content", fill = "Season") +
  theme_minimal()


#log SM graph

ggplot(data = combinedSMlog, aes(x = Block, y = MeanVWC_Log, fill = Season)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  labs(x = "Experimental Block", y = "Log-Transformed Mean Volumetric Water Content", fill = "Season") +
  theme_minimal()
```

Box Plot for Mean VWC by treatment and season

```{r}
ggplot(data = combinedSM_data, aes(x = Treatment, y = MeanVWC, fill = Season)) +
  geom_boxplot() +
  labs(x = "Planting Location", y = "Mean Volumetric Water Content", fill = "Season") +
  theme_minimal()

ggplot(data = combinedSMlog, aes(x = Treatment, y = MeanVWC_Log, fill = Season)) +
  geom_boxplot() +
  labs(x = "Planting Location", y = "Log-Transformed Mean Volumetric Water Content", fill = "Season") +
  theme_minimal()
```

In summery there must have been a rain event or something near when the SM data was collected for Spring 2024

## Germination

total germinated seeds in each season

```{r}
# Combine germination data for all seasons
germination_data <- bind_rows(
  spring24 |> select(Block, Treatment, GerminatedAcorns) |> mutate(Season = "Spring"),
  summer24 |> select(Block, Treatment, GerminatedAcorns) |> mutate(Season = "Summer"),
  fall24 |> select(Block, Treatment, GerminatedAcorns) |> mutate(Season = "Fall")
)

#change the order of the seasons to match daata collection dates 
germination_data <- germination_data |>
  mutate(Season = factor(Season, levels = c("Spring", "Summer", "Fall")))

# Summarize total germinated seeds per season
germination_summary <- germination_data |>
  group_by(Season) |>
  summarise(Total_Germinated = sum(GerminatedAcorns, na.rm = TRUE))

print(germination_summary)

germination_summary |>
  ggplot(aes(x = Season, y = Total_Germinated, fill = Season)) +
  geom_col() +
  labs(title = "Total Germinated Seeds by Season", x = "Season", y = "Total Germinated")

```

### Germinated seeds by block and treatment

```{r}
germination_by_block_treatment <- germination_data |>
  group_by(Block, Treatment, Season) |>
  summarise(Total_Germinated = sum(GerminatedAcorns, na.rm = TRUE))

print(germination_by_block_treatment)

```

### Germination rate

```{r}
# Add Total_Seeds_Planted column 
germination_data <- germination_data |> mutate(Total_Seeds_Planted = 128)  

# Calculate germination rates
germination_data <- germination_data |>
  mutate(Germination_Rate = GerminatedAcorns / Total_Seeds_Planted)

# Summarize germination rates
germination_rate_summary <- germination_data |>
  group_by(Block, Treatment, Season) |>
  summarise(Avg_Germination_Rate = mean(Germination_Rate, na.rm = TRUE))

print(germination_rate_summary)

germination_data |>
  ggplot(aes(x = Treatment, y = Germination_Rate, fill = Season)) +
  geom_boxplot() +
  facet_wrap(~ Block) +
  labs(title = "Germination Rate by Treatment and Block", x = "Treatment", y = "Germination Rate")

```

### Statistical Analysis

#### Effect of Treatment, Block, and Temperature on Germination Rate

```{r}

# Load temperature data
HOBOData24 <- readRDS("HOBOData24.rds")

# Merge germination data with temperature data
germination_temp_data <- merge(germination_data, final_data, by = c("Block", "Treatment"))

# Linear model
germination_model <- lm(Germination_Rate ~ Treatment + Block + Avg_Temperature, data = germination_temp_data)

# Summary of the model
summary(germination_model)

# Diagnostic plots
plot(germination_model)
```

average blocks separate by treatment - effect on treatment

#### Soil Moisture effect on germination rate

```{r}
germ_moist_data <- merge(springSM, germination_data, by = c("Treatment", "Block"))

# Merge soil moisture and germination data

# Check the structure of the merged data
str(germ_moist_data)

# Visualize the relationship between soil moisture and germination rate
germ_moist_data |>
  ggplot(aes(x = MeanVWC, y = Germination_Rate, color = Treatment)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Effect of Soil Moisture on Germination Rate",
    x = "Mean Volumetric Water Content (Soil Moisture)",
    y = "Germination Rate"
  ) +
  facet_wrap(~ Block) 

# Fit a linear model to test the effect of soil moisture on germination rate
lm_germination <- lm(Germination_Rate ~ MeanVWC, data = germ_moist_data)

# Summarize the model
summary(lm_germination)


```

## Soil / Bulk Density

do the soils vary

```{r}

```

do they vary based on micro habitat -\> treatment / block / temp

```{r}

```

block as a random effect -\> how soils vary based on treatment

```{r}

```

block as a fixed effect do soils vary across site -\> notice yes

```{r}

```

controlling for the variation -\> how do they vary across blocks

```{r}

```

\

## Natural Recruitment EDA ?

```{r}

```

## Why we planted more Salvia oop 

canceled lol

```{r}

```

# Predictive Model

```{r}


# Check the levels of each variable
levels(final_data$LifeStage)
levels(final_data$Treatment)
levels(final_data$Block)

final_data <- final_data |>
  mutate(
    LifeStage = as.factor(LifeStage),
    Treatment = as.factor(Treatment),
    Block = as.factor(Block)
  )

# Identify rows with NA levels in categorical variables
invalid_rows <- final_data |>
  filter(is.na(LifeStage) | is.na(Treatment) | is.na(Block))

# Remove invalid rows
final_data <- final_data %>%
  filter(!is.na(LifeStage) & !is.na(Treatment) & !is.na(Block))

# Combine the  datasets
seedling_data <- bind_rows(fall24, spring24, summer24)


# Remove LifeStage from final_data before join
final_data <- final_data |>
  select(-LifeStage) |>
  left_join(seedling_data |> select(Tag, LifeStage), by = "Tag")

# Example: Predicting Growth Rate
model <- lm(Growth_Rate ~ Treatment + Block + LifeStage + Avg_Temperature, data = final_data)
summary(model)
```

acorns vs seedlings across microclimate condtions

tag numbers that need replacing

subset data for replaceiment
